---
title: "Data overview, by strategy, for all protein families"
author: "Sergio Picart-Armada and Angela Lopez del Rio"
date: "28/02/2021"
output:
  html_document:
    toc: true
    theme: united
    toc_float: true
    number_sections: true
    highlight: textmate
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
options(knitr.kable.NA = '')
```

# Load files

```{r}
library(tidyverse)
library(magrittr)
library(corrplot)
library(emmeans)

df.raw <- read.csv("performances/data_count_df.csv")
v.strategy <- levels(df.raw$strategy)

summary(df.raw)

config <- new.env()
source("config.R", local = config)
```

## Overview

Data dimensions:

```{r}
dim(df.raw)
```

Column names:

```{r}
colnames(df.raw)
```


First rows:

```{r}
head(df.raw, 3)
```

Number of rows for each resampling scheme, separated by protein type.
Each square should equal to the number of folds (10):

```{r}
table(df.raw$strategy, df.raw$protein_type)
```

## Clean data frame format

```{r}
# convert the Counter({0: 53086, 1: 49628}) format to actives/inactives
# watchout! sometime counter starts with 1: and then 0:
dict_to_label <- function(x) {
  # look for 1:
  actives <- gsub("(.+1: )([0-9]+)([^0-9].+)", "\\2", x)
  # look for 0:
  inactives <- gsub("(.+0: )([0-9]+)([^0-9].+)", "\\2", x)
  
  paste0(actives, "/", inactives)
}

# dict_to_label(df.raw$training) 

df.clean <- mutate_at(
  df.raw, c("training", "validation", "test"), dict_to_label
) %>%
  select(protein_type, strategy, fold, training, validation, test)
```


```{r}
# check some random values
# 
# gpcr, fold 5, resampling_before_clustering, validation, inactives
# Counter({1: 17324, 0: 17274})
rnd.1 <- subset(
  df.clean, 
  fold == "5" & 
    protein_type == "GPCRs" & 
    strategy == "resampling_before_clustering")$validation
stopifnot(rnd.1 == "17324/17274")

# proteases, fold 8, resampling_before_clustering, training, active: 50987
# Counter({0: 51930, 1: 50987})
rnd.2 <- subset(
  df.clean, 
  fold == "8" & 
    protein_type == "proteases" & 
    strategy == "resampling_before_clustering")$training
stopifnot(rnd.2 == "50987/51930")
```

## Export table

### Full table

```{r}
df.clean
```

The table is interesting, but would be too large to fit in the supplement.
It makes more sense to aggregate over folds.

### Fold-averaged table

```{r}
# average a column
average_character <- function(char, ...) {
  sp <- strsplit(char, "/") 
  actives <- as.numeric(sapply(sp, head, 1))
  inactives <- as.numeric(sapply(sp, tail, 1))
  
  paste0(
    round(mean(actives), ...), 
    "/", 
    round(mean(inactives), ...)
  )
}

average_character(df.clean$training, digits = 1)

df.foldavg <- dplyr::group_by(df.clean, protein_type, strategy) %>%
  summarize_at(c("training", "validation", "test"), average_character)
```

Table:

```{r}
df.foldavg
```

Export to latex:


```{r}
tex.foldavg <-  knitr::kable(
  df.foldavg, format = "latex", 
  caption = "Total number of protein-compound pairs in each strategy and protein family, displayed in the actives/inactives format. Number of actives and inactives were averaged over the 10 folds.", 
  linesep = c("", "", "", "\\addlinespace"), 
  booktabs = TRUE, label = "npairs-avg-strat") %>%
    kableExtra::kable_styling(latex_options = c("scale_down", "hold_position"))

writeLines(tex.foldavg, "10_overview_allfamilies_foldavg.tex")
```


# Reproducibility

```{r}
sessionInfo()
```

