---
title: "Exploring NR performance"
author: "Sergio Picart-Armada and Angela Lopez del Rio"
date: "27/02/2021"
output:
  html_document:
    toc: true
    theme: united
    toc_float: true
    number_sections: true
    highlight: textmate
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
options(knitr.kable.NA = '')
```

## Load files


```{r}
library(tidyverse)
library(magrittr)
library(corrplot)
library(emmeans)

df.raw <- read.csv("performances/ratios_df_rdm_baseline_NR.csv")
v.strategy <- levels(df.raw$strategy)
df.raw.nobaseline <- subset(df.raw, is_baseline == "False")

summary(df.raw)

config <- new.env()
source("config.R", local = config)
```

Data dimensions:

```{r}
dim(df.raw)
```

Column names:

```{r}
colnames(df.raw)
```


First rows:

```{r}
head(df.raw, 3)
```

Number of rows for each resampling scheme, separated by baseline or not:

```{r}
table(df.raw$strategy, df.raw$is_baseline)
```

NAs in training positives ratio?

```{r}
table(df.raw.nobaseline$strategy, is.na(df.raw.nobaseline$ratio_training))
```

```{r}
table(df.raw.nobaseline$strategy, is.finite(df.raw.nobaseline$ratio_test))
```

Resampling after clustering has less proteins in all the folds?
No, it happens the same way in all the folds:

```{r}
table(df.raw.nobaseline$strategy, df.raw.nobaseline$fold, df.raw.nobaseline$is_baseline)
```

Is it always the same proteins?

```{r}
# split(df.raw, df.raw$fold) %>%
#   mapply(VennDiagram::venn.diagram, 
#          x = split(.$Protein, .$strategy), 
#          filename = paste0(head(.$fold, 1), "_venn.png"))
```


Test set: how often do we have proteins with a ratio of 0, or 1? 
Does this explain why resampling after clustering has way less proteins?

```{r}
eps <- 1e-6
table(df.raw.nobaseline$strategy, cut(df.raw.nobaseline$ratio_test, 
                                      breaks = c(-eps, 0, 1-eps, Inf)))
```

Number of familites (should be only one):

```{r}
table(df.raw.nobaseline$family)
```


Summary of metrics by scheme:

```{r}
df.raw.nobaseline %>%
  by(., .$strategy, function(x) summary(x[c("auroc", "f1")]))
```

### Sanity check - duplicated proteins

Are the row numbers unique? Now they are:

```{r}
sum(duplicated(df.raw$X))
```

And in the non-baseline?

```{r}
sum(duplicated(df.raw.nobaseline$X))
```


Check that every protein ID appears at most once in every `strategy` and `fold`.

```{r}
df.raw.nobaseline %>%
    group_by(strategy, fold) %>%
    summarise(max_occurrences = max(table(DeepAffinity.Protein.ID)))
```

There are proteins occurring more than once in every strategy.
Which ones?

```{r}
df.duplicated <- df.raw %>%
    group_by(strategy, fold, is_baseline) %>%
    mutate(is_dup = duplicated(DeepAffinity.Protein.ID)) %>%
    filter(is_dup)
df.duplicated
```

List of unique identifiers:

```{r}
prot.duplicated.id <- unique(df.duplicated$DeepAffinity.Protein.ID) %>% 
  as.character
prot.duplicated.id
```


Are those actually duplicates in all the rows? Apparently not, but the difference is far from obvious

```{r}
dim(df.raw.nobaseline)

select(df.raw.nobaseline, -X) %>% unique %>% dim
```

All the affected rows:

```{r}
subset(df.raw, DeepAffinity.Protein.ID %in% prot.duplicated.id)
```

Decision: take out the second occurrence in each one

```{r}
df.clean <- df.raw %>%
    group_by(strategy, fold, is_baseline) %>%
    mutate(is_dup = duplicated(DeepAffinity.Protein.ID)) %>%
    filter(!is_dup) %>%
    ungroup %>%
    mutate(fold = as.factor(fold), 
           baseline = ifelse(is_baseline == "False", "DL model", "Random baseline"))
```

Sanity check: is the issue fixed now?

```{r}
df.clean %>%
    group_by(strategy, fold, is_baseline) %>%
    summarise(max_occurrences = max(table(DeepAffinity.Protein.ID))) %>%
    extract2("max_occurrences") %>% 
    all(. == 1L) %>%
    stopifnot

df.clean.nobaseline <- subset(df.clean, is_baseline == "False")
df.clean.baseline <- subset(df.clean, is_baseline == "True")
```

Dimensions of final data

```{r}
dim(df.clean)
```

Competitive methods only:

```{r}
dim(df.clean.nobaseline)
```

Save clean table

```{r}
write.csv(df.clean, file = config$file.cleanperf.nr)
```


# Boxplots with metrics overview

### F1

```{r}
ggplot(df.clean, aes(x = strategy, y = f1, fill = strategy)) +
    geom_boxplot() +
    facet_wrap(~baseline) +
    scale_fill_manual(values = config$col.strategy) +
    theme(aspect.ratio = 1, legend.position = "none") +
    config$gg_45
```

### AUC

```{r}
ggplot(df.clean, aes(x = strategy, y = auroc, fill = strategy)) +
    geom_boxplot() +
    facet_wrap(~baseline) +
    scale_fill_manual(values = config$col.strategy) +
    theme(aspect.ratio = 1, legend.position = "none") +
    config$gg_45
```


# Reproducibility

```{r}
sessionInfo()
```

